---
title: "ces21_variable_labels"
author: "Ran"
date: "2025-07-08"
output: html_document
---

## Variable labels
Imported the CES2021 dataset from dta file, convert coded variables to human-readable labels, and extract variable metadata for documentation.

```{r setup, include=FALSE}
library(haven)
library(tidyverse)

# load data
data <- read_dta("dataverse_files/2021 Canadian Election Study v2.0.dta")
```

The original CES 2021 dataset was provided in .dta (Stata) format. Upon loading it into R via the haven package, the variable labels (stored as metadata attributes) exhibited two key issues:

Garbled text or encoding errors — due to incorrect character decoding from .dta format (possibly non-UTF-8 encoding).

Truncated labels — several labels appeared to be cut off mid-sentence, likely due to Stata’s historical 80-character limit for variable labels.

```{r}
# get labels
var_labels <- sapply(data, function(x) {
  lbl <- attr(x, "label")
  if (is.null(lbl)) {
    NA_character_
  } else {
    # Merge multiples into one label
    lbl_clean <- paste(lbl, collapse = " ")
    lbl_clean <- gsub("[\r\n\t]", " ", lbl_clean)
    lbl_clean <- trimws(lbl_clean)
    if (lbl_clean == "") NA_character_ else lbl_clean
  }
})

# create mapping table
var_label_df <- data.frame(
  variable_name = names(data),
  variable_label = var_labels,
  stringsAsFactors = FALSE
)
```

```{r}
# output variable labels

fileConn <- file("variable_labels.csv", open = "wb")
writeBin(charToRaw("\xEF\xBB\xBF"), fileConn)  # UTF-8 BOM
close(fileConn)

write.table(
  var_label_df,
  file = "variable_labels.csv",
  sep = ",",
  row.names = FALSE,
  col.names = TRUE,
  append = TRUE,
  fileEncoding = "UTF-8"
)
```
Check original data file. Confirmed truncated variable labels were caused by source file limitation (Max 80 chars)

```{r}
label <- attr(data[["cps21_not_vote_for_1"]], "label")
nchar(label)
```


